---
title: "The R Package equateMultiple: A Tutorial"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: spacelab
    highlight: pygments
vignette: >
  %\VignetteIndexEntry{The R Package equateMultiple: A Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Functionalities

The equateMultiple package computes:

* __Equating coefficients__ between __multiple forms__.
* __Synthetic item parameters__ (sort of mean of the item parameter estimates from different forms).
* __Standard errors__ of the equating coefficients and the synthetic item parameters.

# Data preparation

Data preparation follows the same steps of the equateIRT package.

Load the package equateMultiple and the data

```{r}
library("equateMultiple")
data("data2pl", package = "equateIRT")
```

Estimate a two parameter logistic model for 5 data sets with the R package ltm

```{r, message=FALSE}
library("ltm")
m1 <- ltm(data2pl[[1]] ~ z1)
m2 <- ltm(data2pl[[2]] ~ z1)
m3 <- ltm(data2pl[[3]] ~ z1)
m4 <- ltm(data2pl[[4]] ~ z1)
m5 <- ltm(data2pl[[5]] ~ z1)
```

Extract the item parameter estimates and the covariance matrices

```{r}
estm1 <- import.ltm(m1, display = FALSE)
estm2 <- import.ltm(m2, display = FALSE)
estm3 <- import.ltm(m3, display = FALSE)
estm4 <- import.ltm(m4, display = FALSE)
estm5 <- import.ltm(m5, display = FALSE)
estm1$coef[1:3, ]
estm1$var[1:3, 1:3]
```

Create a list of coefficients and covariance matrices

```{r}
estc <- list(estm1$coef, estm2$coef, estm3$coef, estm4$coef, estm5$coef)
estv <- list(estm1$var, estm2$var, estm3$var, estm4$var, estm5$var)
test <- paste("test", 1:5, sep = "")
```

Create an object of class `modIRT`

```{r}
mods <- modIRT(coef = estc, var = estv, names = test, display = FALSE)
coef(mods$test1)[1:5]
```

The linkage plan

```{r}
lplan<-linkp(coef = estc)
lplan
```


# Multiple equating coefficients

Estimation of the equating coefficients using
the multiple mean-mean method. Form 1 is the base form.

```{r}
eqMM <- multiec(mods = mods, base = 1, method = "mean-mean")
summary(eqMM)
```

Estimation of the equating coefficients using the 
multiple mean-geometric mean method.

```{r}
eqMGM <- multiec(mods = mods, base = 1, method = "mean-gmean")
summary(eqMGM)
```

Estimation of the equating coefficients using the multiple 
item response function method.

```{r}
eqIRF<-multiec(mods = mods, base = 1, method = "irf")
summary(eqIRF)
```

Estimation of the equating coefficients using the multiple item response function method.
The initial values are the estimates obtained with the 
multiple mean-geometric mean method.

```{r}
eqMGM <- multiec(mods = mods, base = 1, method = "mean-gmean", se = FALSE)
eqIRF<-multiec(mods = mods, base = 1, method = "irf", start = eqMGM)
summary(eqIRF)
```

Estimation of the equating coefficients using the multiple test response function method.

```{r}
eqTRF<-multiec(mods = mods, base = 1, method = "trf")
summary(eqTRF)
```

# Synthetic item parameters

The synthetic item parameters and their standard errors can be
extracted as follows (using the multiple item response function method).

```{r}
eqIRF$as
eqIRF$bs
eqIRF$se.as
eqIRF$se.bs
```

# Equated scores


Equated scores with the true score equating method

```{r}
scTSE<-score(eqIRF)
round(scTSE,3)
```

Equated scores with the observed score equating method, 
avoiding computation of standard errors

```{r}
scOSE<-score(eqIRF, method = "OSE", se = FALSE)
round(scOSE,3)
```

